---
title: "Hypothesis Testing & Effect Size"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Did Community Policing Actually Improve Response Times?

The city of Westbrook implemented a new community policing program in January. The police chief claims the program has improved emergency response times. Before the program, the average response time was **8.2 minutes** with a standard deviation of **1.5 minutes**. You've collected data from 20 randomly selected emergency calls after the program started.

Your editor wants to know two things: Is the improvement real? And is it big enough to matter?

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Set up the data and the hypotheses (2 points)

Before we run any tests, we need to state our hypotheses. The null hypothesis is what we assume to be true until we have evidence otherwise — like "innocent until proven guilty." Fill in the REPLACE_ME values.

**Null Hypothesis (H₀):** there has not been a change in call reponse times. 

**Alternative Hypothesis (H₁):** call response times have improved

```{r}
# Response times (in minutes) after the community policing program
response_times <- c(
  7.8, 8.1, 7.5, 7.9, 8.3,
  7.6, 7.4, 8.0, 7.7, 7.8,
  8.2, 7.3, 7.9, 7.5, 8.1,
  7.6, 7.8, 7.4, 7.7, 7.9
)

# Known baseline (before the program)
baseline_mean <- 8.2
baseline_sd <- 1.5

# Create a tibble
response_data <- tibble(
  call_number = 1:20,
  response_time = response_times
)

response_data
```

### Task 3: Calculate summary statistics (1 point)

Let's see what the new response times look like. Replace the REPLACE_ME values.

```{r}
summary_stats <- response_data |>
  summarize(
    sample_size = n(),
    sample_mean = mean(response_time),
    sample_sd = sd(response_time),
    standard_error = sample_sd / sqrt(sample_size)
  )

summary_stats
```

**Answer this question:** The sample mean is lower than the baseline of 8.2 minutes. How much lower? Does that seem like a meaningful improvement to you, just based on the numbers? What would you want to know before deciding?

It is 0.5 lower than the baseline of 8.2. I think that shows a meaningful improvement because there is such a low number difference. Also, there is only a standard deviation of  0.2 which means that there is little variation between answers. I would want to know if that call time really made a difference with actual response to crime and if that short amount of time generally helped. 

### Task 4: Perform a one-sample t-test (2 points)

Now we formally test the police chief's claim. Since we're testing whether response times *decreased*, we use a one-sided test (`alternative = "less"`). Replace the REPLACE_ME with the baseline mean we're comparing against.

```{r}
# Set our significance level
alpha <- 0.05

# Perform the t-test
t_test_result <- t.test(
  response_data$response_time,
  mu = 8.2,
  alternative = "less"
)

t_test_result
```

### Task 5: Interpret the t-test results (2 points)

Let's pull out the key numbers. Look at the p-value: is it less than our alpha of 0.05?

```{r}
# Extract key values
p_value <- t_test_result$p.value

results <- tibble(
  `Baseline mean` = paste(baseline_mean, "minutes"),
  `New mean` = paste(round(summary_stats$sample_mean, 2), "minutes"),
  `Difference` = paste(round(baseline_mean - summary_stats$sample_mean, 2), "minutes"),
  `p-value` = round(p_value, 6),
  `Significant?` = if_else(p_value < alpha, "Yes - reject null hypothesis", "No - fail to reject")
)

results
```

**Answer this question:** Based on the p-value, can we reject the null hypothesis? What does this tell the police chief — and what does it NOT tell him?

Yes, the p-value is very low, therefore we can reject the null hypothesis. This tells the chief that the original idea that there was no change is wrong. However, it does not tell us that there actually was a huge decrease in the call response time, it just tells us that there was a change. 

### Task 6: Visualize the results

```{r}
ggplot(response_data, aes(x = response_time)) +
  geom_histogram(binwidth = 0.2, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = baseline_mean, color = "red", linetype = "dashed", size = 1.2) +
  geom_vline(xintercept = summary_stats$sample_mean, color = "darkgreen", size = 1.2) +
  annotate("text", x = baseline_mean + 0.05, y = 4,
           label = paste("Baseline:", baseline_mean),
           color = "red", hjust = 0, fontface = "bold") +
  annotate("text", x = summary_stats$sample_mean - 0.05, y = 3.5,
           label = paste("New Mean:", round(summary_stats$sample_mean, 2)),
           color = "darkgreen", hjust = 1, fontface = "bold") +
  labs(
    title = "Emergency Response Times: Before vs. After Community Policing",
    subtitle = paste0("p-value = ", round(p_value, 6)),
    x = "Response Time (minutes)",
    y = "Number of Calls"
  ) +
  theme_minimal()
```

### Task 7: Calculate the effect size (3 points)

The t-test told us the improvement is statistically significant — it's real, not due to chance. But is it *meaningful*? That's what effect size measures. We'll calculate **Cohen's d**, which tells us how large the difference is in standard deviation units.

The formula is: d = (baseline mean - sample mean) / baseline standard deviation

Replace the REPLACE_ME values.

```{r}
effect_size <- (baseline_mean - summary_stats$sample_mean) / baseline_sd

effect_size_interpretation <- case_when(
  abs(effect_size) < 0.2 ~ "Negligible",
  abs(effect_size) < 0.5 ~ "Small",
  abs(effect_size) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `Cohen's d` = round(effect_size, 3),
  `Interpretation` = effect_size_interpretation,
  `Difference in minutes` = round(baseline_mean - summary_stats$sample_mean, 2),
  `Difference in seconds` = round((baseline_mean - summary_stats$sample_mean) * 60, 0)
)
```

**Answer this question:** The result is statistically significant, but the effect size is small. In practical terms, the improvement is about 25 seconds per call. The police chief is claiming success. Would you write a story saying the program works? How would you frame the findings honestly — acknowledging both that the improvement is real AND that it's small?

I do not think I would write an article saying that the program is working 100%. Well there is a decrease in response the data is so small that there really is not much of a difference being made. If I were to write the article I would make sure readers knew that it was a small difference and that even though there was a decrease it was not by much. 

### Task 8: The journalist's framework (5 points)

The videos you watched describe two questions you need to ask about any statistical result:

1. **Is it statistically significant?** (Is the result unlikely to be due to chance?)
2. **Is it meaningfully significant?** (Is the effect large enough to matter?)

Apply this framework to the community policing data:

**Question 1:** Write 1-2 sentences answering whether the result is statistically significant and how you know.

I think that the results are statistically significant. The videos described that something that is statistically significant is something that disproves the null hypothesis and has data to back it. In this example the null hypothesis was rejected and there was a low p-value so this proves that the data is statistically significant. 

**Question 2:** Write 1-2 sentences answering whether the result is meaningfully significant. Consider: the program cost the city $2 million to implement. Is a 25-second improvement in response times worth that investment? What context would you need to answer that question?

I do not think the results are meaningfully significant. The program cost $2 million for only 25 seconds to be cut off from response times. I think that for the money being put in the results were much lower and did not reap as much of a benefit. I do not think the investment was worth it. I would need to know how much money could be used to actually make a differnce and if there was a cheaper way to do this. I would also need to know how long the system has been in place so I could tell if the program has had enough time to be effective. 

**Question 3:** Write a headline and a 2-3 sentence lede for a story about these findings. Your text should accurately reflect both the statistical significance and the effect size — don't let the police chief's framing drive the story if the data tells a more nuanced one.

headline: small improvements in police response time calls into question benefits of new program
lede: After the police department's $2 million new program to improve response time showed success by reducing the call time by 25 seconds. However, while there is data to suggest the program worked, the money put into it caused debate over how successful the project truly was. 



When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
