```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

## Does Poverty Predict Math Scores? Testing a School Board Member's Claim

### Introduction

A school board member in Montgomery County, Pennsylvania has claimed at a public meeting that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests. That's a testable claim, and as data journalists, our job is to test it.

We have school-level data from Montgomery County, Pa. that includes each school's FARMS percentage (the share of students eligible for Free and Reduced-price Meal System) and the percentage of students who met or exceeded math expectations on state tests. We'll use linear regression to evaluate whether the school board member's claim holds up, and we'll learn something important about what happens when you trust a model before examining your data.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_pa.csv")
```

### Task 1: Examine the Data Before You Plot Anything

Before we visualize anything, let's look at what we're working with.

```{r}
summary(schools$`FARMS Pct`)

schools |> arrange(desc(`FARMS Pct`)) |> head(10)

schools |> arrange(`FARMS Pct`) |> head(10)
```

Look at the distribution of FARMS percentages. Do any values look unusual or suspicious? What might it mean if a school has a FARMS percentage of exactly 100? Write two or three sentences describing what you see and whether anything concerns you.

I think that the values with 100 points look suspicious. If a school has that percentage it could mean that someone lied or there are actually students that are getting perfect scores. I think that those scores should be looked into to see if it is accurate or if someone lied. 


### Task 2: Create a Scatterplot and Describe the Results

Now let's visualize the relationship between FARMS percentage and math scores for all schools.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County PA")
```

Describe what you see. Is there a clear trend? Does the regression line look like it represents the data well, or does it seem to be pulled in an odd direction? Connect what you observed in Task 1 to what you see here.

There is not a clear trend. The scattterplot shows that the data is all over the place and none of it meets on the line that goes through the graph. The regression line seems to be random and does not look like it has any meaning to it. This goes with what I said in task 1 because the 100s are alone at the end and all the other data is on the other side of the graph. This shows that these outliers must be an accident and are most likely not accurate. 


### Task 3: Run the Regression Model

Let's build the linear regression model using all schools.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

There are three things to examine: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct.

The **p-value** tells you whether the relationship is statistically significant. The **Adjusted R-squared** tells you how much of the variation in math scores is explained by FARMS percentage — it's already a percentage, so a value of 0.025 means 2.5%, not 25%. The **coefficient** on FARMS Pct tells you the direction and size of the relationship: positive means higher FARMS predicts higher math scores; negative means higher FARMS predicts lower scores.

Report all three values and describe what they tell you. Does this model support the school board member's claim? Is this even a model worth using?

The p-value is 0.12 this is a low number which shows that there is most likely no correlation. The adjusted R-squared is 0.02 which is extremely low and shows that the data is inaccurate. The coeffiecient for FARMS Pct is 0.12 which is also extremely low. I think that the data shows this is not worth using becayse all of the statisitcs show that there is no correlation between the data and the claim. 


### Task 4: What's Driving the Problem? Filter and Rerun.

In Task 1, you likely spotted schools with a FARMS percentage of 100. Those values are suspicious — in most districts, having *every* student on free/reduced lunch is unusual and may indicate a reporting artifact or data error. Let's see what happens when we exclude them.

```{r}
schools_filtered <- schools |> filter(`FARMS Pct` < 100)

model_filtered <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools_filtered)
summary(model_filtered)
```

Now compare the two models side by side:

```{r}
cat("Full dataset model:\n")
cat("  Coefficient:", coef(model)[2], "\n")
cat("  p-value:", summary(model)$coefficients[2,4], "\n")
cat("  Adjusted R-squared:", summary(model)$adj.r.squared, "\n\n")

cat("Filtered dataset model (FARMS < 100):\n")
cat("  Coefficient:", coef(model_filtered)[2], "\n")
cat("  p-value:", summary(model_filtered)$coefficients[2,4], "\n")
cat("  Adjusted R-squared:", summary(model_filtered)$adj.r.squared, "\n")
```

How did removing those schools change the model? Did the coefficient change direction? Did the p-value or R-squared improve meaningfully? What does this tell you about the influence of suspicious data values on a regression model?

Removing the schools changed the model and made the data appear somewhat more reliable. However, the p-value and r-squared value did not change as much as the coefficient which tells me that there is still not much accuracy even when you remove the 100s. This tells me that the suspicious data might not be exactly what is causing the data to be so low. This is because even without it the numbers did not change drastically and are still quite low. Therefore, I think there might be another reason why the data is appearing inaccurate besides the 100s. 


### Task 5: Even the Better Model Has Problems

Even after filtering, the Adjusted R-squared remains very low. This means FARMS percentage — even when the data is clean — explains very little of the variation in math scores across these schools.

Look at the filtered scatterplot:

```{r}
schools_filtered |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage (Excluding 100% FARMS Schools)")
```

With a low R-squared, the residuals from this model aren't very meaningful as labels of "over-performing" or "under-performing" — the model doesn't know enough to make that judgment reliably. But let's calculate them anyway to see the spread:

```{r}
schools_filtered <- schools_filtered |>
  mutate(
    predicted = predict(model_filtered),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools_filtered |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

What are the largest residuals? Given the model's weak R-squared, how much confidence would you have in labeling these schools as over- or under-performers? Why?

The largest residuals are 36-24. This says that there is that much of a difference between the data and what is actually happening. Therefore, I would not have much confidence labeling the schools as over or under because the residual scores are so high. 


### Task 6: Write the Story Pitch

You've now run two models, examined the data quality and seen that even a cleaned-up model explains very little of the variation in math scores. Return to the school board member's claim: that schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch (3-5 sentences) that addresses the following: Does the data support or contradict the claim? What data quality issues did you find? What would you need - different data, a different model, additional reporting - before you could responsibly publish a story about this claim? Who would you call first?

I would write a story about how the data does not appear to show significance and that there are many outliers that show there is odd variannce. I would dspecifically look into why the scores for p-value and r-squared do not change much despite removing the 100 scores that originally seemes to show inconsistency. I would want to look at more data and see if this happens every year in order to see if there are more serious problems or if this is a one time thing. 
