---
title: "SoccerFouls"
---

Do fouls matter in women's college soccer? In this assignment, you'll be working with a dataset that contains information about soccer matches from last season. You'll load the data, perform some basic statistical analysis, and create a visualization to explore the relationship between fouls and the goal differential (and thus the outcome). Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the soccer dataset from https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/refs/heads/main/data/ncaa_womens_soccer_matchstats_2024.csv

```{r}
matches <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/refs/heads/main/data/ncaa_womens_soccer_matchstats_2024.csv")
```


### Task 3: Create a new column called `goal_diff` that calculates the difference between the two teams' goals.

```{r}
matches <- matches |> mutate(goal_diff = team_score - opponent_score)
```



### Task 4: Use `summarize()` and the `cor()` function to calculate the correlation between fouls and goal differential, using the pearson method.

```{r}
matches |> summarize(correlation = cor(fouls, goal_diff, method="pearson"))
```



### Task 5: Create a scatterplot of fouls and goal differential, using `geom_point()`

```{r}

matches |> 
  ggplot() +
  geom_point(aes(x=goal_diff, y=fouls)) +
  geom_smooth(aes(x=goal_diff, y=fouls), method="lm")

```



### Task 6: Describe the results of both the correlation calculation and the scatterplot.

The results of the correlation tell us that there is a low chance that fould impact a teams chance to make goals. The dots on the scatterpoint show that there is not often times a relation. In some cases the teams that won ended up having the most amount of fouls.Therefore both the correlation calculation and the scatterplot show that there is not an impact. 

#Task 7: Choose another variable and repeat the steps in Task 4 and 5, then describe. 


```{r}
matches |> summarize(correlation = cor(corners, goal_diff, method="pearson"))
```

```{r}

matches |> 
  ggplot() +
  geom_point(aes(x=goal_diff, y=corners)) +
  geom_smooth(aes(x=goal_diff, y=corners), method="lm")

```
The data sows that there is potentially a impact of corners in relation to points. The correlation calculation is much stronger which shiws that there is a high chance tgat the number of corners impacts a game. Also, the scatterplot is heavier on the right side which suggests that the more corners means more goals.  

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
