---
title: "HW10: Significance Testing & Effect Size"
name: REPLACE WITH YOUR NAME
---

You previously calculated the mean, margin of error and confidence interval for monthly car thefts in Prince George's County. Your editor now wants to know: it _looks_ like car thefts are decreasing lately, but is that change real, and how big is it? In this assignment, you'll use significance testing and effect size to answer those questions. Load the tidyverse and the data to get started.

### Task 1: Load the tidyverse and the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}

# Read the data
car_thefts_pgcounty <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") 
car_thefts_pgcounty|>
  arrange(month)  # Sort by date

```

### Task 2: Split the data and test for significance (4 points)

We'll split the monthly data into two halves, and then ask whether there's a statistically significant difference between these periods. To do that, we'll use a t-test with our alpha at 0.05. First, fill in the hypotheses, then replace the REPLACE_ME values in the code.

**Null Hypothesis:** car thefts are not decreasing

**Alternative Hypothesis:** car thefts are decreasing

```{r}

# Find the halfway point and add a column placing each month in the first or second halves

midpoint <- ceiling(nrow(car_thefts) / 2)
car_thefts <- car_thefts |>
  mutate(period = if_else(row_number() <= midpoint, "First Half", "Second Half"))

# Perform t-test comparing the two time periods
t_test_periods <- t.test(total ~ period, data = car_thefts)

# Display results
t_test_periods

```

**Answer this question:** What is the p-value? Is it less than 0.05? Can we reject the null hypothesis? What does this tell your editor?

The p-value is 2.14 which is much higher than 0.05. Yes, we can reject the null hypothesis based on this data becuase there is such a high p-value which shows the confidence that the data is significant. This tells my editor that the data does not suggest that there is a severe enough decrease in car thefts for the data to hold meaning. 

### Task 3: Visualize the comparison (4 points)

Let's see the two periods side by side with their confidence intervals. Replace the REPLACE_ME values.

```{r}
# Summarize by period
period_summary <- car_thefts |>
  group_by(period) |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts,
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

period_summary |>
ggplot(aes(x = period, y = mean_thefts, fill = period)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_text(aes(label = round(mean_thefts, 1)), vjust = -0.5, size = 4) +
  labs(
    title = "Comparing Car Thefts: First Half vs. Second Half",
    subtitle = paste0("p-value = ", round(t_test_periods$p.value, 5)),
    x = "Time Period",
    y = "Average Number of Car Thefts"
  ) +
  scale_fill_manual(values = c("First Half" = "#66c2a5", "Second Half" = "#fc8d62")) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Answer this question:** Look at the error bars (confidence intervals) on the chart. Do they overlap? What does that tell you about whether the difference between the two periods is real?

No they do not overlap. This tells me that there is a decrease potentially because there is enough of a drop that the confidence data does not include anything in the original first half. This means that in the second half there was lower crime rates whereas in the first half there were more higher numbers.This tells us that there could be a difference since there is no overlap, however I think more details would be needed to really decide. 

### Task 4: Calculate the effect size (5 points)

We've established that the decline is statistically significant. But how large is it? Let's calculate Cohen's d to measure the effect size. For a two-sample comparison, we use the pooled standard deviation:

**Pooled SD** = sqrt(((n1 - 1) × sd1² + (n2 - 1) × sd2²) / (n1 + n2 - 2))

**Cohen's d** = (mean1 - mean2) / pooled SD

Replace the REPLACE_ME values. You'll need to pull the values from `period_summary`.

```{r}
# Extract the values from period_summary
first_half_stats <- period_summary |> filter(period == "First Half")
second_half_stats <- period_summary |> filter(period == "Second Half")

# Calculate pooled standard deviation
pooled_sd <- sqrt(
  ((first_half_stats$months - 1) * first_half_stats$sd_thefts^2 +
   (second_half_stats$months - 1) * second_half_stats$sd_thefts^2) /
  (first_half_stats$months + second_half_stats$months - 1)
)

# Calculate Cohen's d
cohens_d <- (first_half_stats$mean_thefts - second_half_stats$mean_thefts) / pooled_sd

# Interpret it
effect_label <- case_when(
  abs(cohens_d) < 0.2 ~ "Negligible",
  abs(cohens_d) < 0.5 ~ "Small",
  abs(cohens_d) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `First half mean` = round(first_half_stats$mean_thefts, 1),
  `Second half mean` = round(second_half_stats$mean_thefts, 1),
  `Difference` = round(first_half_stats$mean_thefts - second_half_stats$mean_thefts, 1),
  `Cohen's d` = round(cohens_d, 3),
  `Interpretation` = effect_label
)
```

**Answer this question:** How would you describe this effect size? Compare it to the effect sizes we calculated in class (the response times activity had a small effect, the housing prices activity had a medium effect). What does a large effect size mean for the story you'd write about car thefts?

I would describe this effect size as large. In class the effect sizes we calculated were lower in number or showed not as much of a significant change. A large effect size shows me that there was an actually large enough difference between the two data points. I think that I would have more confidence saying that the results are accurate. 

### Task 5: Putting it all together (7 points)

You now have three pieces of evidence: the confidence intervals from last week, the significance test from this assignment, and the effect size.

1. In Part 1, your editor asked "just tell me if car thefts are going up or down". Now you have significance testing. Write 2-3 sentences telling your editor the answer, using all three pieces of evidence (CI, p-value, and effect size).

Based off the Cohen d test it is clear that there is an actually large difference between the data and this effect size tells us that the difference between the two is large enough for the number to have meaning. Also, the confidence intervals do not overlap which means that there is a difference between the two and they do not coincide with each other. There is a large enough gap that the confidence intervals are far apart which tells us that there is an even more likelihood that the data has meaning. 

2. Write a headline and a 2-3 sentence lede for a story about car thefts in Prince George's County. Unlike the response times story from class (where the effect was small and the headline needed to be cautious), this is a case where the data strongly supports a clear conclusion. Your writing should reflect that — but still note the limitations (correlation vs. causation, what else might explain the decline).

Headline: car thefts in Prince George's County are decreasing in number 

Lede: In Prince George's County car theft rates are decreasing as shown by the effect size which tells you how much of a increase or decrease there really was between original and existing data. The data suggests very highly that there is a decrease in car thefts, however it should be notes that correlation does not equal causation which could still skew these results. 

When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
