---
title: "HW7b - Probability and the Normal Distribution"
---

In this assignment, you'll build on your z-score skills to answer probability questions using the normal distribution. You'll return to the White House salaries data and use it to answer the kinds of "what are the chances?" questions that come up in journalism all the time. This sets the stage for sampling, which we'll cover next.

### Task 1: Load the tidyverse and the data (1 point)

Load the tidyverse and the White House salaries data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv into a dataframe called `wh_salaries`

```{r}
library(tidyverse)
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```

### Task 2: Filter and summarize (2 points)

You may remember from HW3 that some White House staffers have a salary of $0. Let's filter those out and focus on paid employees, saving the result to a new dataframe called `wh_paid`. You need to calculate the mean, median and standard deviation of the salary column using `wh_paid`:

```{r}
wh_paid <- wh_salaries |>
  filter(SALARY > 0)
wh_paid <- wh_salaries |>  
  mutate(
    mean_salaries = mean(SALARY),
    sd_salaries = sd(SALARY),
    median_salaries  = median(SALARY)
  )

```

Now create a histogram of the paid salaries with a vertical line for the mean and two more for ±1 standard deviation, similar to what you did in HW6 with Spotify data.

```{r}
wh_paid |>
  ggplot() +
  geom_histogram(aes(x=SALARY), binwidth = 10000) +
  geom_vline(aes(xintercept = mean(SALARY)), color = 'red', linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(SALARY) - sd(SALARY)), color = 'blue', linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(SALARY) + sd(SALARY)), color = 'blue', linetype = "dashed", size =1)

```

**Describe the shape of the distribution. Is it roughly normal? Skewed? Where do most salaries cluster?**

The distribution is not normal. It is skewed to the left because most of the data falls on the left side of the graph. Also, it is not normal because there are two outliers on the far left and far right which makes it so the curve is going up and down throughout the distribution. The majority of the salaries actually do not occur where the mean line was produced showing that the data is skewed as well and not normally distributed. 

### Task 3: The Empirical Rule (3 points)

The empirical rule (also called the 68-95-99.7 rule) says that for a normal distribution:
- About 68% of data falls within 1 standard deviation of the mean
- About 95% falls within 2 standard deviations
- About 99.7% falls within 3 standard deviations

Let's test this. We'll calculate the salary ranges for 1, 2, and 3 standard deviations from the mean, then count how many employees actually fall in each range.

```{r}
mean_sal <- mean(wh_paid$SALARY)
sd_sal <- sd(wh_paid$SALARY)
total_n <- nrow(wh_paid)

empirical <- tibble(
  range = c("Within 1 SD", "Within 2 SD", "Within 3 SD"),
  lower = c(mean_sal - sd_sal, mean_sal - 2 * sd_sal, mean_sal - 3 * sd_sal),
  upper = c(mean_sal + sd_sal, mean_sal + 2 * sd_sal, mean_sal + 3 * sd_sal),
  expected_pct = c(68, 95, 99.7),
  actual_count = c(
    sum(wh_paid$SALARY >= (mean_sal - sd_sal) & wh_paid$SALARY <= (mean_sal + sd_sal)),
    sum(wh_paid$SALARY >= (mean_sal - 2 * sd_sal) & wh_paid$SALARY <= (mean_sal + 2 * sd_sal)),
    sum(wh_paid$SALARY >= (mean_sal - 3 * sd_sal) & wh_paid$SALARY <= (mean_sal + 3 * sd_sal))
  ),
  actual_pct = round((actual_count / total_n) * 100, 1)
)

empirical
```

**How closely does the data follow the empirical rule? What does any deviation tell you about the distribution?**

The data follows the empiracal rile decently closely. The actual percentage of the first SD was 65.8% while it is expected to be 68% so it was only about 2% away from what it is expected to be based on the empirical rule. However, the second SD is decently far off because the percentage was actually 89.4% and it is meant to be 95% this is almost a 6% difference which is pretty large. However, within 3 SD the percentage was only 0.3% off which is very accurate. This tells me that the second SD is where the data begins to skew and not be as accurate as it should be based on the empirical rule. 

### Task 4: From z-scores to probability with pnorm() (4 points)

Remember the car thefts z-score activity from class? Well, I want to return to that briefly, introduce a new tool called pnorm() and then apply it here. So open that CarTheftZScores notebook and, at the bottom, copy and paste in the contents here: https://gist.githubusercontent.com/dwillis/25102889f576b1707576b85e0be0c353/raw/1ca2be9636497cb5293ff7b1f4e6ed8e27a9e049/car_theft_pnorm.rmd

Re-run all of the blocks in CarTheftZScores notebook, including the code you just pasted in. What's happening there is you are using a function called `pnorm()` to find the probability of a value being less than or equal to a certain z-score. The basic formula to find that probability is:

```
1 - pnorm(z_score)
```

Where `z_score` is one row's z-score. Now apply the same approach to salaries.

Your editor asks: **"What's the probability that a randomly selected White House employee makes more than $150,000?"**

To answer this:
1. Calculate the z-score for $150,000
2. Use `pnorm()` to find the probability. `pnorm()` gives you the area under the normal curve to the LEFT of a z-score — that is, the probability of being BELOW that value.
3. Subtract from 1 to get the probability of being ABOVE $150,000
4. Check your answer against the actual data

```{r}

salary <- 150000
z_score <- (salary - mean_sal) / sd_sal

prob_less <- pnorm(z_score)
prob_more <- 1 - prob_less
prob_more

# Check against actual data
actual_prob <- sum(wh_paid$SALARY > 150000) / total_n
actual_prob
```

**Write a sentence describing this probability in plain English. How does the normal approximation (prob_less) compare to the actual percentage?**

This probability says that there is a low likelihood because the closer to 1 the number is the more it is to be certain. 0.17 tells us that there is around a 17% likelihood of the event occuring in mostly both scenarios, since there is only a slight difference between the two. The normal aproximation compares to the actual percentage because it is a lot lower since the actual percentages are closer to the normal empirical rule numbers wheras the normal approximation is a lot lower. 

### Task 5: Probability between two values (3 points)

Now your editor asks: **"What percentage of White House employees make between $80,000 and $120,000?"**

You'll need to calculate z-scores for those two figures, use `pnorm()` for each, and subtract to find the probability between them. Update the REPLACE_ME figures with the appropriate numbers.

```{r}

z_lower <- (salary - mean_sal) / sd_sal
z_upper <- (salary - mean_sal) / sd_sal

prob_between <- pnorm(z_upper) - pnorm(z_lower)
prob_between

# Verify with actual data
actual_between <- sum(wh_paid$SALARY >= 80000 & wh_paid$SALARY <= 120000) / total_n
actual_between

```

**Interpret this result. What does this "middle range" tell us about the White House salary structure?**

This result tells us that around 30% of White House employees make between $80,000 to $120,000. This middle range tells us that mroe employees either make less than this number or much more. This tells us that the whitehouse does not have its salaries structured in a way that would make sense to most people. Less people are making the more "average" salaries and more people are making either extremely low salaries or extremely high salaries. 

### Task 6: Expected vs. observed (3 points)

Let's compare what the normal distribution *predicts* to what we actually *see*. This idea — comparing expected to observed — is one you'll use again later in the course.

Create these salary bins and calculate the expected count (from the normal distribution) and the observed count (from the actual data) for each:

```{r}
bins <- tibble(
  category = c("Below $60k", "$60k-$90k", "$90k-$120k", "$120k-$150k", "Above $150k"),
  lower = c(-Inf, 60000, 90000, 120000, 150000),
  upper = c(60000, 90000, 120000, 150000, Inf)
) |>
  mutate(
    expected_pct = pnorm(upper, mean = mean_sal, sd = sd_sal) - pnorm(lower, mean = mean_sal, sd = sd_sal),
    expected_count = round(expected_pct * total_n, 1),
    observed_count = c(
      sum(wh_paid$SALARY < 60000),
      sum(wh_paid$SALARY >= 60000 & wh_paid$SALARY < 90000),
      sum(wh_paid$SALARY >= 90000 & wh_paid$SALARY < 120000),
      sum(wh_paid$SALARY >= 120000 & wh_paid$SALARY < 150000),
      sum(wh_paid$SALARY >= 150000)
    ),
    difference = observed_count - expected_count
  )

bins |> select(category, expected_count, observed_count, difference)
```

**Where does reality differ most from what the normal distribution predicts? What might explain those differences?**

Reality differs a lot compared to what the the expected result would be. In the above $150k section there was the lowest difference which made sense. There was an extremely high difference un the range of 60-90% which is also much higher than it is supposed to be. I also thought it was very interesting that in the middle two categories the numbers were much lower than they were supposed to be. 

### Task 7: Story pitch (4 points)

Based on your probability analysis, write a headline and lede for a story about White House salaries. Your lede should incorporate at least one probability or percentage you calculated.

Then write 2-3 sentences describing how you would use probability thinking to advance this story — what could you compare these probabilities to, and what additional analysis would help?

Story idea: I think it would be interesting to look into why exactly there is such a large gap between the White House salaries and what would be expected based on the empirical rule. I think it would be interesting to look more into why this happens and see if there is an underlying reason. 

Headline: Inconsistent White House salary's cause debate over root cause. 

Lede: In the White House, the amount of employees receiving a middle of the raod rage, between $90k and $120k is 26% lower than expected while 69% of employees are making less, causing questions about why this is occurring. 

I would use probability thinking in this story to show what is the expected salary and then compare that middle of the road salary with how much money these people are actually making. I would compare the probability of someone making the middle of the road salary and the lower salary to show that someone is more likely to make less event though this goes against the empiric rule. 



### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
