---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse


```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
MandM_data <- read_csv('https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv')
```

### Task 3: Sample 5 rows of data from the dataset


```{r}
MandM_data |>
  slice(10,11,12,13,14)
```


### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms


```{r}
MandM_data |>
  slice(10,11,12,13,14) |>
  summarise(mean_sample = mean(red), sd_sample = sd(red))
MandM_data |>
  slice(10,11,12,13,14) |>
  summarise(mean_sample2 = mean(black), sd_sample2 = sd(black))
```


### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4


```{r}
MandM_data |>  
  slice(10,11,12,13,14) |>
  summarize(mean_total = mean(total), sd_total = sd(total))
```

### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

The mean and sd between the red and black M and M's did not have much of a difference. The mean and sd in the sample of the red M and M's was slightly smaller than the sample of the black M and M's in my sample. However, the number was very low and between the two means there was only a 0.9 difference and between the sd there was only a 1 point difference. Both mean and sd for both colors were very far away from the mean and sd of the population. There was almost a 50% difference in each of the colors from the total. Therefore, I have inferred that the amount of each color in my sample was similar because there is almost a 50-50 spread between the two colors. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

In the data, as mentioned before, my sample size had almost an equal amount of red and black M and M's in total. The difference between the mean and sd for the two colors were almost the same and the total mean and sd were about 50% of either the red or black M and M's. Also, when I added the amount of each color I found that there was 345 red M and M's in my sample and 348 M and M's in the black sample. Meaning there was only a difference of three M and M's which is why my data was so evenly spit. This tells me that sometimes the data can work like this where there is a random almost perfect split. However, I know that oftentimes this is not accurate based on the other M and M activity that we did. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
