---
title: "HW9: Car Theft Confidence Intervals & Margin of Error"
name: Lena LaJoy
---

Your editor wants to know what a "typical" month of car thefts looks like in Prince George's County. Not just the average — she wants to know how confident you can be in that number. Using monthly figures, confidence intervals and margin of error, you'll come up with an answer and decide whether the data supports a clear story. As with the previous homework, provide answers in this notebook when prompted and you will have to fill in values where you see REPLACE_ME. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# LOAD THE TIDYVERSE
```

### Task 2: Load the Prince George's county crime data

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  arrange(month)  # Sort by date

# Basic inspection of the data
glimpse(car_thefts)

```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any trend over time. What's your initial reaction to the change in car thefts? Is there a trend? If your editor asked you "how many car thefts happen in a typical month?", what would you say based on this chart alone?

My initial reaction is that car thefts have decreased pretty significantly lower throughout the years. Also, there appears to be a spike in thefts around the summer months and a decrease in thefts around the winter months. Based on this chart, I would say that it depends on the season and then I would say that on a typical summer month there is around 300-400 thefts whereas in a winter month thefts go down to 300-200. 

```{r}

car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2026)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

```

### Task 4: Calculate the mean, standard error and margin of error

We've got a small number of months, so we need to quantify the uncertainty around our average. We'll calculate the standard error (which measures how much the sample mean could vary), and the margin of error (which gives us the plus/minus range). Replace the REPLACE_ME values based on what we did in class.

```{r}

theft_summary <- car_thefts |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(total),
    margin_error = qt(0.975, df = months - 1) * se_thefts
  )

# Display the results
theft_summary

```

**Answer this question:** Explain in your own words what the margin of error tells us here. If you were writing a story, how would you use this number? For example, how would you complete this sentence: "Prince George's County averages about ___ car thefts per month, plus or minus ___."

Here the margin of error tells us in this case how the entire total of thefts compares to what the specific sample that is being looked at is. It compares the standard error with a statistic to check for how accurate those are. I would use this to explain how the numbers are accurate and shows the confidence in the standard error. I would use the data to show accuracy of the data. I would say: Prince George's County averages about 375.6 car thefts per month, plus or minus 94. 


# Task 5: Build the confidence interval

Now let's turn the margin of error into a full confidence interval — the range where we're 95% confident the true average falls. Replace the REPLACE_ME values.

```{r}

theft_summary <- theft_summary |>
  mutate(
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

theft_summary |>
  select(mean_thefts, margin_error, ci_lower, ci_upper)

```

**Answer this question:** If a county official claimed "car thefts are down to about 300 per month," does your confidence interval support that claim? What about a claim of 350 per month? Explain how the confidence interval helps you evaluate each claim.

I would say that this is inaccurate because there is at least 60 more for each month than is mentioned in that claim. I think that a claim of 350 per month is more accurate. However, I think that the best wording would be to say that car thefts are down by about 360 per month because this gives a much more accurate idea of what the data is. The confidence interval helps evaluate because it tells you if the data you are given is actually accurate or if it is skewed. If the confidence intervals show the data being innacurate than the interval worked and you know to maybe not use the data or preference that before publishing it. 

### Task 6: Visualize the data with the confidence interval

Let's make a chart showing the data with the mean and the 95% confidence interval. Does this change anything about your answer from Task 3?

This changes not much about my initial quess. I think it is interesting that the confidence interval is so in line with the mean. I think this shows that the average is accurate and that the extremes are happening when I first said. I think that the confidemce interval is showing where the most frequent thefts are happening. Then the sharp ups and downs show when there are insconsistancies and it lines up with what I originally said about the seasons. 

```{r}
# Make a chart with the mean and CI
car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper),
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month),
           y = theft_summary$mean_thefts + 10,
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)),
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()

```

### Task 7: The margin of error and sample size

In class we learned that MoE = 1/√n × 100 as a quick approximation for polls. That same principle applies here: more data means a smaller margin of error. Let's see what would happen if we had 40 months of data instead of the number we have, assuming the same mean and standard deviation. Replace the REPLACE_ME value with 40.

```{r}

hypothetical_summary <- tibble(
  mean_thefts = theft_summary$mean_thefts,
  sd_thefts = theft_summary$sd_thefts,
  months = 40,
  se_thefts = sd_thefts / sqrt(months),
  margin_error = qt(0.975, df = months - 1) * se_thefts,
  ci_lower = mean_thefts - margin_error,
  ci_upper = mean_thefts + margin_error
)

# Compare the two
comparison <- tibble(
  scenario = c("Actual Data", "40 months"),
  months = c(theft_summary$months[1], hypothetical_summary$months[1]),
  margin_error = c(theft_summary$margin_error[1], hypothetical_summary$margin_error[1]),
  ci_lower = c(theft_summary$ci_lower[1], hypothetical_summary$ci_lower[1]),
  ci_upper = c(theft_summary$ci_upper[1], hypothetical_summary$ci_upper[1])
)

comparison

```

**Answer this question:** How did the margin of error and confidence interval change? What does this tell you about the tradeoff between having more data and being able to make precise claims? If you were covering crime trends, how many months of data would you want before writing a definitive story?

The margin of error changed a lot between the two data sets. This shows that if you were to do more months the chance for an error in accuracy goes up by almost 23% which is a large jump. However, there is not as much of a change in the lower and upper confidence intervals. I was surprised by this because the margin of error changed so much. 

### Task 8: Reporting with uncertainty

Based on everything you've calculated, answer the following:

1. Write a headline and a one-paragraph lede about car thefts in Prince George's County. Your text should properly reflect the uncertainty in the data — don't overstate or understate what the numbers show.

2. Your editor says "Just tell me if car thefts are going up or down." Using only the confidence interval (no significance testing yet), what's the most accurate answer you can give? What would you need to do next to answer that question more definitively?

Headline: 'Car thefts decreased 400 incidents since 2024'

Lede: In Prince George's County, a steady decrease in car thefts occurred since 2024 with spikes in winter and summer months. There is a large differnce between confidence scores for each month suggesting that there is uncertainty in the statistics. 

2. I would tell them that there is a range of around 360-380 where the data could potentially be around 360 lower or 380 higher than it actually is. Meaning that the data might not be accurate because there is rush a large range so I am unable to clearly say if there was a decrease actually or not. 

When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
