---
title: "Confidence Intervals with Sleep Data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Are Students Getting Enough Sleep?

The CDC recommends that adults aged 18-25 get at least 7 hours of sleep per night. A campus health reporter wants to know: are students at this university getting enough sleep?

Remember providing the hours you slept the night before our first class? Your job is to analyze this data and determine what we can confidently say about sleep habits across the entire student body.

This is the core challenge of confidence intervals: using a **sample** to make claims about a **population**.

### Task 1: Load the tidyverse and the data

```{r}
library(tidyverse)

sleep_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/sleep_data.csv")

head(sleep_data)
```

### Task 2: Calculate descriptive statistics

Let's start with what we know about these students. Calculate the mean, standard deviation and number of observations.

```{r}
sleep_stats <- sleep_data |>
  summarise(
    mean_hours = mean(hours),
    sd_hours = sd(hours),
    n = n()
  )

sleep_stats
```

**Answer this question:** Based on these numbers alone, what would you tell the health reporter? Why might it be risky to report the average as the definitive answer for all students?

It would be risky ti report this because while the avergae is slightly above the amount of sleep students are meant to be getting, there is a small sample size. If the tens of thousands of students enrolled only 37 people answered this poll, therefore the data is not accurate because such a small group was asked. I would tell the health reporter that in this specific class the average was 7.7, however this is a small margin compared to everyone in the school. 

### Task 3: Calculate the standard error

The standard error tells us how much our sample mean would vary if we took many different samples of 37 students. It's the key to moving from "this is what 37 students told us" to "this is what we think is true for the whole campus." The formula is: standard deviation divided by the square root of the sample size.

```{r}
sleep_stats <- sleep_stats |>
  mutate(
    se_hours = sd_hours / sqrt(37)
  )

sleep_stats
```

**Answer this question:** The standard error is much smaller than the standard deviation. Why? What does each one measure?

The standard error is much smaller because it is saying that there would not be much of a variation in the sample mean if the sample size was extended and the average was assumed for a larger sample. The standard error only measures the changes within the mean which would cause there to appear to be less changes while the standard deviation looks at the actual difference between the entire data set.  

### Task 4: Calculate a 95% confidence interval

Now we'll build our confidence interval. For a 95% CI with a small sample, we use the t-distribution. The `qt(0.975, df = n - 1)` function gives us the critical value — the number of standard errors we go above and below the mean. We multiply critical value by the standard error to get the margin of error, then add/subtract from the mean.

```{r}
sleep_stats <- sleep_stats |>
  mutate(
    t_critical = qt(0.975, df = n - 1),
    margin_error = t_critical * se_hours,
    ci_lower = mean_hours - se_hours,
    ci_upper = mean_hours + se_hours
  )

sleep_stats |>
  select(mean_hours, margin_error, ci_lower, ci_upper)
```

**Answer this question:** Write one sentence explaining what this confidence interval means in plain English that a reader would understand.

A confidence interval is the range of where someone believes that there would be around a certain amount of accurracy between the results and if it was moved into a bigger scale. 

### Task 5: Visualize the distribution with the confidence interval

This chart shows the histogram of individual sleep hours, plus the mean (red line) and the 95% confidence interval (blue shaded region). The shaded region is the range where we think the true average for all students falls.

```{r}
ggplot(sleep_data, aes(x = hours)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = sleep_stats$mean_hours, color = "red", linewidth = 1) +
  geom_vline(xintercept = sleep_stats$ci_lower, color = "blue", linetype = "dashed") +
  geom_vline(xintercept = sleep_stats$ci_upper, color = "blue", linetype = "dashed") +
  annotate("rect",
           xmin = sleep_stats$ci_lower,
           xmax = sleep_stats$ci_upper,
           ymin = 0, ymax = Inf,
           fill = "blue", alpha = 0.1) +
  annotate("text", x = sleep_stats$mean_hours + 0.1, y = 8,
           label = paste0("Mean: ", round(sleep_stats$mean_hours, 2)),
           hjust = 0, color = "red", fontface = "bold") +
  labs(
    title = "Distribution of Student Sleep Hours",
    subtitle = paste0("95% Confidence Interval for the mean: [",
                     round(sleep_stats$ci_lower, 2), ", ",
                     round(sleep_stats$ci_upper, 2), "]"),
    x = "Hours of Sleep",
    y = "Number of Students"
  ) +
  theme_minimal()
```

**Answer this question:** Look at the chart. Notice that the confidence interval is much narrower than the spread of individual data points. Why? What's the difference between the variation in individual sleep hours and the uncertainty about the average?

The confidence interval is much more narrow because that is the range that would be more likely to be accurate for a larger group of people and shows that it is likely for people to commonly have that narrow answer. The difference between and variation and the uncertainty is that the confidence interval is looking more at the changes in the mean instead of the overal data which causes it to be closer to the mean and a smaller number in general. 

### Task 6: The CDC question — is 7 hours within the interval?

The CDC recommends at least 7 hours. Based on the confidence interval, can we say students are getting enough sleep? Let's check.

```{r}
cdc_recommendation <- 7

sleep_stats |>
  mutate(
    meets_cdc = if_else(ci_lower >= cdc_recommendation, "Yes - entire CI is at or above 7", "No - CI includes values below 7")
  ) |>
  select(mean_hours, ci_lower, ci_upper, meets_cdc)
```

**Answer this question:** Based on this result, which of these headlines is justified? Why?

  a) "Students Average More Than 7 Hours of Sleep"
  b) "Students May Not Be Getting Enough Sleep, Survey Suggests"
  c) "Survey: Students Getting Recommended Amount of Sleep"

The headline that is justified is 3 because it clarifies that the data is from a survey. This is better than option 1 because it says students which makes it seem like everyone is getting 7 hours of sleep while the study is only looking at 37 students. Option 2 cannot be justified because the data shows that students are getting enough sleep even if it is a small sample so this would be misleading. Option 3 is the best because it clarifies that it is not the entire population and it also shows that the students in that sample appear to be receiving the right amount of sleep which is accurate. 

### Task 7: What about a bigger sample?

The margin of error depends on the sample size. What if we had surveyed 100 students instead of 37 and got the same mean and standard deviation? Let's recalculate with a hypothetical sample size 100.

```{r}
hypothetical_n <- 100

hypothetical_stats <- tibble(
  mean_hours = sleep_stats$mean_hours,
  sd_hours = sleep_stats$sd_hours,
  n = hypothetical_n,
  se_hours = sd_hours / sqrt(hypothetical_n),
  t_critical = qt(0.975, df = hypothetical_n - 1),
  margin_error = t_critical * se_hours,
  ci_lower = mean_hours - margin_error,
  ci_upper = mean_hours + margin_error
)

# Compare the two intervals side by side
comparison <- tibble(
  sample_size = c(sleep_stats$n, hypothetical_n),
  margin_error = c(sleep_stats$margin_error, hypothetical_stats$margin_error),
  ci_lower = c(sleep_stats$ci_lower, hypothetical_stats$ci_lower),
  ci_upper = c(sleep_stats$ci_upper, hypothetical_stats$ci_upper)
)

comparison
```

**Answer this question:** How did the margin of error change? What does this tell you about the relationship between sample size and certainty?

With a larger sample size the margun of error decreased by around 0.2%. This tells me that thew relationship between the sample and certainty is higher because just changing how many people are asked directly changes the margin of error which shows how much of a decrease or increase there could be between data. 

### Task 8: Journalistic interpretation

You now have all the statistics you need. Based on what you've calculated:

Write a headline and a one-paragraph lede for a campus news story about student sleep habits. Your text should accurately reflect the confidence interval — don't overstate or understate the findings.

Headline: 
Survey shows some students receive the recommended 7 hours of sleep

Lede: 
A group of students in Journalism Statisitcs were asked how much sleep they received and the averagge answer between these students was slightly above the CDC's recommended seven hours of sleep. However, the data only sampled 37 students, therefore the data should not be expanded to all students because this would alter the data. 
